### 2019-10-11 Manuscript/manuscript.tex
progress throughout project including the development of paleogeo recons using WK07 polygons

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a}). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=4in]{Figures/Tectonic_history.pdf}
\caption{\small{\textbf{Simplified summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}
\label{fig:tectonic_history}
\end{figure}

@@

@@

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean provinces. There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation. A lingering question raised in \citet{Hoffman198a} that still lingers is whether the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and remamalgamation. Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/Laurentia_poles_combined.pdf}
\caption{\small{\textbf{Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard). Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}}}
\label{fig:Laurentia_poles}
\end{figure}

Following Laurentia's amalgamation, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015a} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the following rotations:

@@

Through the Proterozoic there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse. The implied paleolatitudes for an interior point on Laurentia is shown in Figure \ref{fig:Laurentia_paleolatitude}. There are many high-quality paleomagnetic poles in the early Mesoproterozoic with 12  poles between ca. 1480 and 1380 from Laurentia (including Greenland poles). The best-constrained portion of the record is associated with 1110 to 1083 Ma magmatism within the Midcontinent Rift and southwest Laurentia that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins. Overall the record of paleomagnetic poles has a lot of consistency with progressive paths and agreement between poles.

#A nice touch would be to add Paleoproterozoic, Meso, Neo and Phanero onto this figure
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/Laurentia_paleolatitude.pdf}
\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79ºN, lon=92.10ºW). The paleomagnetic poles are compiled in Table \ref{tab:Laurentia_poles}. Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
\label{fig:Laurentia_paleolatitude}
\end{figure}

\section{Paleogeographic reconstructions}

Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:recons_time}. These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's growth. As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block. Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.

The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.

\bibliographystyle{gsabull}
\small\bibliography{../references/allrefs}

>>>>

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a}). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=4in]{../Figures/Tectonic_history.pdf}
\caption{\small{\textbf{Simplified summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}
\label{fig:tectonic_history}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
\caption{\small{\textbf{Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard). Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}}}
\label{fig:Laurentia_poles}
\end{figure}

@@

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean provinces. There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation. Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and remamalgamation. An analysis of the record strengthened with a new paleomagnetic pole from the Slave province in \cite{Mitchell2014a}, supports significant differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga. Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.

@@

Following Laurentia's amalgamation, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015a} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the  rotations shown in Table \ref{tab:terrane_rotations}.

Through the Proterozoic there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse. To visualize the temoporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia is shown in Figure \ref{fig:Laurentia_paleolatitude}. There are many high-quality paleomagnetic poles in the early Mesoproterozoic with 12  poles between ca. 1480 and 1380 from Laurentia (including Greenland poles). The best-constrained portion of the record is associated with 1110 to 1083 Ma magmatism within the Midcontinent Rift and southwest Laurentia that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins. Overall the record of paleomagnetic poles has a lot of consistency with progressive paths and agreement between poles.

%A nice touch would be to add Paleoproterozoic, Meso, Neo and Phanero onto this figure
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79ºN, lon=92.10ºW). The paleomagnetic poles are compiled in Table \ref{tab:Laurentia_poles}. Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
\label{fig:Laurentia_paleolatitude}
\end{figure}

\section*{Paleogeographic reconstructions}

Seeking to developing comprehensive continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types. Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles. Paleomagnetic poles compilations themselves are evolving with better data and improved geochronology. Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\small{\textbf{Paleogeogrpahic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data. These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.}}}
\label{fig:Laurentia_reconstructions}
\end{figure}

 There is an overall lack of published continuous models in the literature for the Proterozoic that can be compared to the compilation of paleomagnetic poles presented herein. The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters. With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g. \citealp{Torsvik2012a}).

 An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017a} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}.  The extent to which the implied position of Laurentia in \cite{Merdith2017a} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and the researchers need to balance varying constraints. The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. go to moderately high southerly latitudes ca. 1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole despite its `A' Nordic rating. The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.

There are more published models that show snapshots at given time intervals \cite[e.g.]{Li2008a}.  The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.

In this work, I develop a continuous paleogeogrpahic reconstruciton of Laurentia constrained the paleomagnetic poles compilation. This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (900-800 Ma for example). One could use interpretations of paleogeographic connections with other cratons (e.g. Baltica in the Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself. The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}. Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:Laurentia_reconstructions.pdf}. These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's growth. As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block. Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}. In Figure \ref{fig:Laurentia_reconstructions.pdf}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.

\section*{Acknowledgements}

Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein. Particular acknowledgement goes to David Evans for maintaining and distributing the compiled pole lists as well as additional efforts of Lauri Pesonen in maintining the Paleomagia database \citep{Veikkolainen2014a}. GPlates, and in particular the pyGPlates API, was utilized in this work \citep{Muller2018b}. Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive python environment \citep{Perez2007a}. This work was supported by NSF CAREER Grant 1847277 awarded to N.L.S.-H.

\bibliographystyle{gsabull}
\small\bibliography{../../references/allrefs}

@@
